---
title: "Data Science Process with COVID 19"
author: "J. Buchanan"
date: "2023-10-05"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(lubridate)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Loading the Data

The first step is to import the raw data from the John Hopkins repository on GitHub.

```{r John Hopkins COVID 19 Data, echo=FALSE}
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in, file_names)

us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])

kable(global_cases[1:5,])
kable(us_cases[1:5,])
```

We will also bring in some global population data that will be used later.

```{r, echo=FALSE}
uid_lkp_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lkp_url) %>% select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))

kable(uid[1:5,])
```

## Cleaning the Data

For this analysis, pivoting on the date columns and turning them into a single column will make viewing the data set much easier. Also, the latitude and longitude columns will not be used in this analysis, so they are removed.

```{r}
global_cases <- global_cases %>% pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long), names_to = "date", values_to = "cases") %>% select(-c(Lat, Long))
global_deaths <- global_deaths %>% pivot_longer(cols = -c('Province/State','Country/Region', Lat, Long), names_to = "date", values_to = "deaths") %>% select(-c(Lat, Long))

us_cases <- us_cases %>% pivot_longer(cols = -(UID:Combined_Key), names_to = "date", values_to = 'cases') %>% select(-c(Lat, Long_))
us_deaths <- us_deaths %>% pivot_longer(cols = -(UID:Population), names_to = "date", values_to = 'deaths') %>% select(-c(Lat, Long_))

kable(us_cases[1:5,])
```

We can take a few more steps now to clean up the data set. To further simplify, we can combine the tables with deaths and cases to view the two columns side by side. Additionally, We can improve visibility in the date columns with the lubridate package. Finally, the tables specific to the United States have several columns that won't be used for this analysis. We can take the opportunity to remove them.

```{r Final US Dataset, echo=FALSE}
global <- global_cases %>% full_join(global_deaths) %>% mutate(date = mdy(date))  %>% rename(Country_Region = 'Country/Region', Province_State = 'Province/State')
us <- us_cases %>% full_join(us_deaths) %>% select(Admin2:deaths) %>% mutate(date = mdy(date))

kable(us[1:5,])
```

The last bit of cleaning will be adding the columns that are missing in the global data, but are present in the US states data. Here, we are simply creating a new combined key by uniting the province and country columns.

```{r echo=FALSE}
global <- global %>% unite("Combined_Key", c(Province_State, Country_Region), sep = ", ", na.rm = TRUE, remove = FALSE)

kable(global[1:5,])
```

Finally, the global population data we imported will be added to the global table.

```{r Final Global Dataset, echo=FALSE}
global <- global %>% left_join(uid, by = c("Province_State", "Country_Region")) %>% select(-c(UID, FIPS)) %>% select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)

kable(global[1:5,])
```

## Data Analysis

The data for the US is currently held at the most granular level it is available from the source. To compare the US data to the global data, the number of cases and deaths will be aggregated up, first by the state, then by the country. This will give various levels to compare the US to the rest of the globe.

```{r US Totals by Date, echo=FALSE}
us_by_state <- us %>% group_by(Province_State, Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths*1000000/Population) %>% select(Province_State, Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

us_totals <- us_by_state %>% group_by(Country_Region, date) %>% summarize(cases = sum(cases), deaths = sum(deaths), Population = sum(Population)) %>% mutate(deaths_per_mill = deaths*1000000/Population) %>% select(Country_Region, date, cases, deaths, deaths_per_mill, Population) %>% ungroup()

us_totals %>% filter(cases > 0) %>% ggplot(aes(x = date, y = cases)) + geom_line(aes(color = "cases")) + geom_point(aes(color = "cases")) + geom_line(aes(y = deaths, color = "deaths")) + geom_point(aes(y = deaths, color = "deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
```

Over time, the number of new cases has leveled off dramatically. To assess the to current impact, we can include a new column that will include the just the new cases each day.

```{r Daily US Cases, echo=FALSE}
us_by_state <- us_by_state %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))
us_totals <- us_totals %>% mutate(new_cases = cases - lag(cases), new_deaths = deaths - lag(deaths))

us_totals %>% ggplot(aes(x = date, y = new_cases)) + geom_line(aes(color = "new_cases")) + geom_point(aes(color = "new_cases")) + geom_line(aes(y = new_deaths, color = "new_deaths")) + geom_point(aes(y = new_deaths, color = "new_deaths")) + scale_y_log10() + theme(legend.position = "bottom", axis.text.x = element_text(angle = 90))
```

From this plot, we can see that the number of new cases is increasingly spread apart. apart from long term trends, there are roughly 1 week periodic spikes in the data,indicating that there are certain days of the week where more new cases are reported.

In addition to the timeline analysis, we can also analyze COVID cases by region. To see which regions were most impacted, we can aggregate the data by state in the US data and by country in the global data.

```{r 10 Regions with most COVID-19 Cases}
us_state_totals <- us_by_state %>% group_by(Province_State) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases/population, deaths_per_thou = 1000*deaths/population) %>% rename(region = 'Province_State') %>% filter(cases > 0, population > 0)

global_nation_totals <- global %>% group_by(Country_Region) %>% summarize(deaths = max(deaths), cases = max(cases), population = max(Population), cases_per_thou = 1000*cases/population, deaths_per_thou = 1000*deaths/population) %>% rename(region = 'Country_Region') %>% filter(cases > 0, population > 0)

global_totals <- us_state_totals %>% full_join(global_nation_totals) %>% filter(region != 'US')

kable(global_totals %>% slice_max(deaths_per_thou, n = 10))
```

```{r 10 Regions with least COVID-19 Cases}
kable(global_totals %>% slice_min(deaths_per_thou, n = 10))
```

After looking through the regions with the minimum and maximum number of deaths per thousand, we can look to see if there is a relationship between the density of cases and the death density for each region. I will create a simple linear model to find the parameters of a potential relationship between these data.

```{r Cases per Thousand vs Deaths per Thousand, echo=FALSE}
mod <- lm(deaths_per_thou ~ cases_per_thou, data = global_totals)

summary(mod)

global_totals_w_pred <- global_totals %>% mutate(pred = predict(mod))

global_totals_w_pred %>% ggplot() + geom_point(aes(x = cases_per_thou, y = deaths_per_thou, color = "data")) + geom_point(aes(x = cases_per_thou, y = pred, color = "prediction"))
```

From the plot, it would appear that a linear relationship is present. However, the data has a very wide variance, with few data points being near the prediction line. Further analysis would be needed to understand if there is any other factors that could help predict death rates.

## Conclusion

Though the COVID-19 data was predictable early on after the onset of the pandemic, over time many variables lead to led to less orderly data. Before performing this analysis, I would have expected the population to be a significant determining factor in the impact of the virus on a certain region. However, in viewing the individual states of the US against the other nations of the world, I found them relatively similar in the impact of the virus on the population. American states were also condensed toward the end of the table with higher cases per thousand. In contrast, the nations that experienced the least impact are in regions that have low international trade and tourism, they are regions I would suspect of having large areas with low population density. This leads me to believe that there are very important social, political, and economic factors that would need to be considered to fully understand the impact COVID-19 had on a given region.

The expectations I had for the data were likely a result of my perspective as a resident of a heavily populated area. Additionally, I was in a position to social distance as much as possible, as well as receive medical attention and vaccines when they were ready. These points have a significant affect on the way I view the impact of the COVID-19 virus. 